---
title: "Pv3R Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pv3R Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

```{r setup}
library(Pv3Rs)
```

<!-- adapted from Inference_example_ICL_3212.R -->
## Synthetic example

We first look at an synthetic example with 3 markers (named `m1`,`m2`,`m3`) and three episodes, with known allele frequencies. 
```{r}
y <- list(enroll = list(m1=c('C','G','T'), m2=c('A','C'), m3=c('C','G','T')),
          recur1 = list(m1=c('C','T'), m2=c('A'), m3=c('A','C')),
          recur2 = list(m1=c('T'), m2=c('A'), m3=c('A')))
fs <- list(m1=c(A=0.27, C=0.35, G=0.18, T=0.20),
           m2=c(A=0.78, C=0.14, G=0.07, T=0.01),
           m3=c(A=0.21, C=0.45, G=0.26, T=0.08))
```
Given the number of alleles observed during each episode, the most parsimonious explanation is that there are three, two, and one genotypes across the three episodes respectively. When performing inference on the recurrence states, the bulk of the computational time lies in computing the log-likelihood of each possible relationship graph over the six genotypes.
```{r, cache=TRUE}
post <- compute_posterior(y, fs, return.RG=TRUE)
```
Note that we did not specify the prior, so a uniform prior across all three recurrence states is assumed by default. We set `return.RG=TRUE` for relationship graphs to be returned. This is by default `FALSE` to conserve memory.

The marginal posterior probabilities of each recurrence state is stored in `post$marg`:
```{r}
post$marg
```
We can access the joint posterior through `post$joint`. Here, we find the most likely sequence of recurrence states:
```{r}
post$joint[which.max(post$joint)]
```
The log-likelihoods of each relationship graph is also returned in the output. Here, we plot the relationship graph for which the data are most likely:
```{r}
RG <- post$RGs[[which.max(sapply(post$RGs, function(RG) RG$logp))]]
# default genotype names
gs <- paste0("g", 1:6)
ts_per_gs <- c(1, 1, 1, 2, 2, 3)
par(mar = c(0.5, 0.5, 0.5, 0.5))
plot_RG(RG_to_igraph(RG, gs, ts_per_gs), edge.curved=0.25, vertex.size=20)
seqs_comp_MLE_RG <- compatible_rstrs(RG, split(gs, ts_per_gs))
```

In this example, the most likely recurrence-state sequence given the data is incompatible with the graph for which the data are most likely (the graph for which the data are most likely is compatible with `r seqs_comp_MLE_RG`). This apparent inconsistency can be understood as follows. It is not due to bias induced by the uniform prior on graphs (see `?compute_posterior` examples). Indeed, when data from the first recurrence are analysed alone, the most probable cause of recurrence is reinfection, but the graph for which the data are most likely is incompatible with reinfection. 

Although sequences `r seqs_comp_MLE_RG` are associated with the graph for which the data are most likely, they are also associated with many graphs for which the data are unlikely or impossible under the model. Meanwhile, the most likely sequence `r names(post$joint[which.max(post$joint)])` is associated with fewer graphs, over which the average conditional probability of the data is larger (see histograms below). The average conditional probability given graphs compatible with a specific sequence is equal to the probability of the data given said sequence (the likelihood of a sequence) because the probability of a graph given a sequence is uniform over graphs compatible with said sequence. Moreover, in the current case where the prior on sequences is uniform, the likelihood is proportional to the prior and the maximum posterior estimate is equal to the maximum likelihood estimate (see scatter plot below). 

```{r, fig.height=10, fig.width=7}
# Extract sequences
seqs <- names(post$joint)

# Extract likelihoods
prob_data_given_graph <- exp(sapply(post$RGs, function(RG) RG$logp))

# For each graph, extract compatible sequences and store in a list
gs_per_ts <- split(gs, ts_per_gs)
seq_per_RG_list <- lapply(post$RGs, compatible_rstrs, gs_per_ts)

# For each sequence, return TRUE if graph ID compatible 
seq_per_RG_matx <- sapply(seqs, function(seq) sapply(seq_per_RG_list, function(x) seq %in% x))

# For each sequence with non-zero posterior probability, plot histogram of likelihoods
poss_seqs <- names(sort(post$joint[post$joint > 0], decreasing = T))
likelihoods <- array(0, dim = c(1, length(post$joint)), dimnames = list(NULL, names(post$joint)))

par(mfrow = c(3, 2))
ML <- max(prob_data_given_graph)
xlims <- range(prob_data_given_graph)
for(poss_seq in poss_seqs) {
  hist(prob_data_given_graph[seq_per_RG_matx[,poss_seq]], breaks = 100, 
       xlim = xlims, xlab = "Likelihood", col = "cornflowerblue", border = "cornflowerblue", 
       main = sprintf("%s (compatible with %s relationship graphs)", poss_seq, 
                      sum(seq_per_RG_matx[,poss_seq])))
  likelihoods[1,poss_seq] <- mean(prob_data_given_graph[seq_per_RG_matx[,poss_seq]])
  abline(v = likelihoods[1,poss_seq], col = "blue")
}
```

```{r, fig.height=5, fig.width=7}
par(mfrow = c(1, 1))
plot(y = post$joint, x = likelihoods[1,], ylab = "Posterior", xlab = "Likelihood")
lmfit <- lm(post$joint ~ likelihoods[1,])
abline(a = 0, b = lmfit$coefficients[2], col = "blue")
text(y = post$joint[post$joint > 0], x = likelihoods[1,post$joint > 0], pos = 1, 
     labels = names(post$joint[post$joint >0]))
```

<!-- adapted from Inference_example_ICL_3212.R -->
## Microsatellite data

Next, we apply our methods to microsatellite data, which is available [here](https://github.com/jwatowatson/RecurrentVivax/blob/master/RData/GeneticModel/MS_data_PooledAnalysis.RData).
```{r}
load('../data/MS_data_PooledAnalysis.RData')
MSs_all <- tail(names(MS_pooled), 8) # marker names
head(MS_pooled, 10) # first 10 rows
```
For this example, we use a Empirical Bayes estimate of allele frequencies inferred from the alleles observed in the enrolment episodes.
```{r}
D_weight_Prior <- 1 # uniform Dirichlet prior for allele frequencies
ind_enrol = which(MS_pooled$Episode==1) # Indices of enrolment episodes
Alpha_Posteriors <- apply(MS_pooled[,MSs_all], 2, function(x, ind_enrol){
  xmax <-  max(x,na.rm=T) # number of possible alleles for this marker
  # prior concentration parameters
  param_vector <- array(D_weight_Prior, dim=xmax, dimnames=list(1:xmax))
  # observed data summarised as counts
  obs_counts <- table(x[ind_enrol])
  # posterior concentration parameters
  param_vector[names(obs_counts)] <-  param_vector[names(obs_counts)] + obs_counts
  return(param_vector)
}, ind_enrol)
# use posterior mean as Empirical Bayes estimate
Fs_Combined <- sapply(Alpha_Posteriors, function(x){x/sum(x)})
```

We group rows of the full dataset by patient ID, and filter out cases that are not currently suitable for inference
```{r, results="hold"}
library(plyr)
y.dfs <- dlply(MS_pooled, 'ID')
no_recur <- c()
too_many_genos <- c()
na_only <- c()
wrong_order <- c()
suitable <- c()

for(i in 1:length(y.dfs)) {
  y.df <- y.dfs[[i]]
  
  # no recurrence, skip
  if(length(unique(y.df$Episode_Identifier)) == 1) {
    no_recur <- c(no_recur, i)
    next
  }
  
  # >8 genotypes considered too complex
  if(nrow(y.df) > 8) {
    too_many_genos <- c(too_many_genos, i)
    next
  }
  
  # only use markers for which there is at least one non-NA
  ms <- MSs_all[apply(!is.na(y.df[MSs_all]), 2, any)]
  if(length(ms) == 0) {
    na_only <- c(na_only, i)
    next
  }
  
  # ensure that Episode is non-decreasing
  if(!all(order(y.df$Episode) == 1:nrow(y.df))) {
    wrong_order <- c(wrong_order, i)
    next
  }
  
  suitable <- c(suitable, i)
}
writeLines(paste("No recurrences:", length(no_recur)))
writeLines(paste("More than 8 genotypes (too complex):", length(too_many_genos)))
writeLines(paste("NAs only:", length(na_only)))
writeLines(paste("Episode numbers not in increasing order:", length(wrong_order)))
writeLines(paste("Can proceed with inference:", length(suitable)))
```
Using a uniform prior over recurrence states, we perform inference on the recurrence states for the first 5 cases:
```{r}
results <- list()
for(i in suitable[1:5]) {
  y.df <- y.dfs[[i]]
  patient.id <- names(y.dfs)[i]
  writeLines(paste("ID:", patient.id))

  # only use markers for which there is at least one non-NA
  ms <- MSs_all[apply(!is.na(y.df[MSs_all]), 2, any)]

  # transform data frame format to format taken by 'compute_posterior'
  y.by.episode <- dlply(y.df, 'Episode_Identifier')
  y <- lapply(y.by.episode, function(episode) {
    setNames(lapply(ms, function(m) {
      alleles <- episode[m][!is.na(episode[m])] # extract non-NAs
      if(length(alleles) > 0) return(alleles)
      return(NA) # if all are NAs, change empty vector to NA
      }), ms)
    })

  results[[patient.id]] <- compute_posterior(y, Fs_Combined)
}
```
