---
title: "Intra-episode siblings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intra-episode siblings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%", # Line width taken up by plot
  fig.dpi = 300, # For quality 
  fig.width = 7, # Increase to zoom out
  fig.height = 2, 
  cache = TRUE
)
```

```{r setup, echo = FALSE}
library(Pv3Rs)
#options(warn = -1) # Suppress warnings about missing and limited data
sm <- function(...) suppressMessages(...) # Rename to shorten function call
```

When MOIs exceed two, Pv3Rs sums over graphs with cliques of three or more
intra-episode siblings. However, groups of three or more not-half intra-episode
siblings collapse to groups of two when modelled MOIs are based on estimates
from bulk (not single-cell) parasite genetic data because

- MOI estimates based on maximum per-marker allele counts count at most two for
groups of three or more not-half siblings because groups of not-half siblings 
can contribute at most two alleles per marker.

- MOI estimates based on heteroallelic marker counts (e.g., those estimated by
TheRealMcCOIL, coiaf and MOIRE [AT-to-add-REFS]) count at most two for groups of
three or more not-half siblings because data from a group of not-half siblings
can be at-most as diverse across markers as the two parental genotypes from
which they draw (check).

Although Pv3Rs sums over graphs with cliques of three or more intra-episode
siblings, when the data input of `compute_posterior` are as diverse as the
modelled MOIs, graphs with cliques of three or more intra-sibling parasites have
zero likelihood; when modelled MOIs are more diverse than the data input of
`compute_posterior` (e.g., because the user specifies elavated MOIs), graphs
with cliques of three or more intra-episode siblings have non-zero liklihoods.
In either case, provided modelled MOIs are based on estimates from per-episode
parasite genetic data analysed in bulk (not single-cell) the true number of
siblings per group collapses to two per group; see following example.

## Example of a group of four intra-episode siblings collapsing to two

Before showing how groups of intra-episode siblings collapse to pairs of
intra-episode siblings, we write a function to simulate data of a given allelic
richness (marker cardinality) on an enrolment episode comprising a stranger plus
a group of four siblings, two from one oocyst, two from another, all drawing
from the same two unrelated parental genotypes, and a recurrence with one
sibling. Technically, the enrolment episode contains five genetically distinct
genotypes and thus has a MOI of five.

```{r}
simulate_data <- function(marker_cardinality){
  
  # Magic numbers / quantities
  set.seed(5) # For reproducibility
  n_markers <- 200 # Number of markers
  n_strangers <- 3 # Number of stranger parasites
  n_oocysts <- 2 # Number of oocysts to draw from
  
  # Derived quantities
  alleles <- letters[1:marker_cardinality]
  markers <- paste0("m", 1:n_markers) # Marker names
  
  # Uniform allele frequencies 
  fs <- sapply(markers, simplify = FALSE, 
               function(m) setNames(rep(1/marker_cardinality, marker_cardinality), alleles))
  
  # Sample strangers
  strangers <- sapply(1:n_strangers, function(i) {
    sapply(markers, function(t) sample(names(fs[[t]]), size = 1, prob = fs[[t]]))
  })
  
  # Designate strangers
  parents <- strangers[, 1:2]
  
  # Map the markers to chromosomes. Assume equally sized chromosomes — reasonable
  # if and only if we later assume an equal number of crossovers per chromosome
  chrs_per_marker <- round(seq(0.51, 14.5, length.out = n_markers))
  
  # Sample parental allocations dependently per-oocyst 
  cs <- lapply(1:n_oocysts, function(o) recombine_parent_ids(chrs_per_marker))
  
  # Construct children from parental allocations 
  all_children <- lapply(1:n_oocysts, function(o) {
    oocyst_chidren <- sapply(1:n_markers, function(i) {
      sapply(1:ncol(cs[[o]]), function(j) parents[i,cs[[o]][i,j]])
    })
    colnames(oocyst_chidren) <- markers
    return(oocyst_chidren)
  })
  
  # Make enrolment infection
  enrol <- apply(rbind(all_children[[1]][1:2,], 
                       all_children[[2]][1:2,], 
                       strangers[,3]), 2, unique, simplify = F)
  
  # Make paired data
  data <- list(enrol = enrol, recur = as.list(all_children[[1]][1,]))
  
  return(list(data = data, fs = fs))
}
```
However, MOI estimates based on maximum per-marker allele counts are 
three and one when markers are polyallelic:

```{r}
polyallelic <- simulate_data(10) 
determine_MOIs(polyallelic$data)
```

And one and two when markers are biallelic: 

```{r}
biallelic <- simulate_data(2) 
determine_MOIs(biallelic$data)
```

Suppose we estimate MOIs of 3 and 1 for the biallelic data using an external
software that exploits heteroallelic marker counts (rather than the maximum
number of per-marker alleles), and input into `compute_posterior` these external
estimates. Providing data are simulated on a large number of markers (200 above),
we recover almost exactly the same posterior probabilities using both the

- polyallelic data without user-specified MOIs
- biallelic data with user-specified MOIs three and one

```{r echo = FALSE}
poly <- sm(compute_posterior(y = polyallelic$data, 
                              fs = polyallelic$fs, return.logp=T))
bi <- sm(compute_posterior(y = biallelic$data, MOIs = c(3,1),
                           fs = biallelic$fs, return.logp=T))

poly$marg
bi$marg
```
However, given polyallelic data the likelihoods of graphs with cliques of three
intra-episode siblings are zero: 

```{r echo = FALSE, fig.cap="Graphs with zero likelihood"} 
MOIs <- c(3,1) # MOIs based on data 
gs <- paste0("g", 1:sum(MOIs)) # Genotype names
ts_per_gs <- rep(1:length(MOIs), MOIs) # Episode indices per genotype
llikes <- sapply(poly$RGs, function(RG) RG$logp) # Extract log likelihoods
par(mar = c(0, 0.5, 0, 0.5), mfrow = c(1,5)) # Plotting parameters
for(RG in poly$RGs[which(is.infinite(llikes))]) {
plot_RG(RG_to_igraph(RG, gs, ts_per_gs),
        vertex_palette = "Greys",
        edge.curved = 0.08,
        vertex.size = 40,
        edge.width = 3,
        vertex.frame.width = 2,
        vertex.label.cex = 1,
        vertex.label.color = "black")
}
```

Whereas, given biallelic data, all relationships have non-zero likelihoods:

```{r} 
llikes <- sapply(bi$RGs, function(RG) RG$logp) # Extract log likelihoods
any(is.infinite(llikes)) # Are there any minus infinity log likelihoods?
```

# Graphs with cliques or three or more siblings elevate upper bounds 

```{r}
all_MOIs <- sapply(colnames(maxima), function(x) as.numeric(strsplit(x, split = "")[[1]]))
maxMOIs <- sapply(all_MOIs, function(x) max(x))
```

The summation over graphs with cliques or three or more siblings is possibly
redundant given these graphs have no practical support. Even when these graphs
have zero likelihood, they contribute to maximum probabilities of recrudescence
/ reinfection through the conditionally uniform prior on graphs. However, the
contribution is generally very small:

```{r}
# Compare with and without summation
plot(NULL, xlim = c(0.5, 1), ylim = c(0.5,1),
     ylab = "without summation", xlab = "with summation"); abline(a = 0, b = 1)
points(x = maxima["theory_I_with", maxMOIs > 2], 
       y = maxima["theory_I_wout", maxMOIs > 2], pch = 21, bg = "red")
diffs <- maxima["theory_I_with", maxMOIs > 2] - maxima["theory_I_wout", maxMOIs > 2]
title(main = sprintf("Largest difference: %s", 
                     round(max(abs(diffs), na.rm = TRUE), 4))) # remove na.rm once maxima regenerated
```

```{r}
diffs <- maxima["theory_C_with", maxMOIs > 2] - 
  maxima["theory_C_wout", maxMOIs > 2]

# Compare with and without summation
plot(NULL, xlim = c(0.5, 1), ylim = c(0.5,1),
     ylab = "without summation", xlab = "with summation"); abline(a = 0, b = 1)
points(x = maxima["theory_C_with", maxMOIs > 2][abs(diffs) > 0.015], 
       y = maxima["theory_C_wout", maxMOIs > 2][abs(diffs) > 0.015],
       pch = 1, cex = 2)
points(x = maxima["theory_C_with", maxMOIs > 2], 
       y = maxima["theory_C_wout", maxMOIs > 2], pch = 21, bg = "yellow")

title(main = sprintf("Largest difference: %s", 
                     round(max(abs(diffs), na.rm = TRUE), 4))) # remove na.rm once maxima regenerated

text(x = maxima["theory_C_with", maxMOIs > 2][abs(diffs) > 0.015], 
       y = maxima["theory_C_wout", maxMOIs > 2][abs(diffs) > 0.015], 
     labels = names(all_MOIs)[maxMOIs > 2][abs(diffs) > 0.015], 
     pos = c(1,3), cex = 0.3, offset = 1)
```
The largest absolute difference is less than 0.03 and only five graphs have an absolute difference greater than 0.015. 

