---
title: "Understanding model behaviour"
output: rmarkdown::html_vignette
  toc: true
vignette: >
  %\VignetteIndexEntry{model-ouput}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.dpi=300, # For quality 
  fig.width=7
)
```

```{r setup}
library(Pv3Rs)
```

## Introduction

In this vignette we demonstrate how the statistical model underpinning 
`Pv3Rs::compute_posterior` behaves when it is well specified and misspecified. 

* Posterior output when the model is well specified
  + As a function of marker informativeness
  + As a function of multiplicities of infection (MOIs)
  + When there are no data

* Posterior output when the model is misspecified because of 
  + Genotyping errors or *de novo* mutations
  + Meiotic siblings
  + Parent child-like siblings
  + Half siblings

We have not yet characterised model misspecification due to inbred parasites. 

## Well-specified model behaviour

### Missing / limited data

When data are entirely missing, `compute_posterior` returns the prior. 

```{r}
fs = list(m1 = c("A" = 0.5, "B" = 0.5)) # Allele frequencies
y = list(enrol = list(m1 = NA), recur = list(m1 = NA)) # Data
suppressMessages(compute_posterior(y, fs))$marg # Posterior
```

When data are missing but the user provides MOIs that are incompatible with
recrudescence, `compute_posterior` returns the prior re-weighted to the
exclusion of recrudescence.

```{r}
fs = list(m1 = c("A" = 0.5, "B" = 0.5)) # Allele frequencies
y = list(enrol = list(m1 = NA), recur = list(m1 = NA)) # Data
suppressMessages(compute_posterior(y, fs, MOIs = c(1,2)))$marg # Posterior
```

When genetic data are uninformative, `compute_posterior` returns the prior. 

```{r}
fs = list(m1 = c("A" = 1, "B" = 0)) # Allele frequencies
y = list(list(m1 = "A"), recur = list(m1 = "A")) # Data
suppressMessages(compute_posterior(y, fs))$marg # Posterior
suppressMessages(compute_posterior(y, fs, MOIs = c(1,2)))$marg # Posterior
```

When a single allele repeat is common, all states are possible and the posterior
is close to the prior.

```{r}
fs = list(m1 = c("A" = 0.95, "B" = 0.05)) # Allele frequencies
y = list(enrol = list(m1 = "A"), recur = list(m1 = "A")) # Data
suppressMessages(compute_posterior(y, fs))$marg # Posterior
```

#### Data on only one episode

When only one episode has data and its MOI is one, `compute_posterior` returns
the prior (possibly re-weighted to the exclusion of recrudescence, depending on
the MOIs of the other episodes).

```{r}
# Allele frequencies
fs = list(m1 = c("A" = 0.5, "B" = 0.5))

# Data for different numbers of recurrences
y <- list(enroll = list(m1 = NA), 
          recur1 = list(m1 = "A"), 
          recur2 = list(m1 = NA), 
          recur3 = list(m1 = NA))

# Marginal posterior probabilities given MOIs > 1 for episodes without data
suppressMessages(compute_posterior(y, fs, MOIs = c(3,1,1,2)))$marg
```

When only one episode has data and its MOI is greater than one, the output of
`compute_posterior` depends on the hetero/homoallelic nature of the data. 

If the data are heteroallelic, the posterior remains close to the prior, both
when the MOI increases with and without the number of alleles observed.

Explanation: A heteroallelic call is compatible with intra-infection siblings
and strangers. As such, a heteroallelic call does not constrain the summation
over relationship graphs. However, for relationship graphs with intra-episode
siblings, the heteroallelic data limits summation over identity-by-descent
partitions to partitions with at least two cells for the episode with data. The
lower bound on the cell count increases when the heteroallelic call is
multiallelic.

```{r}
# Allele frequencies
fs = list(m1 = c('A' = 0.2, 'B' = 0.2, 'C' = 0.2, 'D' = 0.2, 'E' = 0.2)) 

# MOI increases with number of observed alleles 
ys <- list(list(enroll = list(m1 = c('A','B')), recur = list(m1 = NA)), # Data 
           list(enroll = list(m1 = c('A','B','C')), recur = list(m1 = NA)), 
           list(enroll = list(m1 = c('A','B','C','D')), recur = list(m1 = NA)), 
           list(enroll = list(m1 = c('A','B','C','D','E')), recur = list(m1 = NA)))
do.call(rbind, lapply(ys, function(y) # Posterior 
  suppressMessages(compute_posterior(y, fs)$marg))) 

# MOI increases due to external input
y <- list(enroll = list(m1 = c('A','B')), recur = list(m1 = NA)) # Data 
MOIs <- list(c(3,1), c(2,2), c(3,2)) # MOIs
do.call(rbind, lapply(MOIs, function(x) # Posterior 
  suppressMessages(compute_posterior(y, fs, MOIs = x)$marg))) 
```

If the data are homoallelic and the allele is rare, the posterior departs from
the prior because intra-episode rare allele repeats favour relationship graphs
with intra-episode siblings.

```{r}
fs = list(m1 = c("A" = 0.001, "B" = 0.999)) # Allele frequencies
y <- list(enroll = list(m1 = c('A')), recur = list(m1 = NA)) # Data 
MOIs <- list(c(2,1), c(3,1), c(2,2), c(3,2), c(3,3)) # MOIs
results <- do.call(rbind, lapply(MOIs, function(x) #  Posterior for different MOIs
  suppressMessages(compute_posterior(y, fs, MOIs = x)$marg)))

# Plot prior departure
par(mar = c(0,0,0,0))
plot_simplex(c("Recrudescence", "Relapse", "Reinfection"), 0.5) # Plot simplex
xy <- apply(results, 1, project2D) # Project probabilities
points(x = xy["x", ], y = xy["y", ], pch = 20, cex = 1, col = 1:length(MOIs))
legend("left", col = 1:length(MOIs), pch = 20, title = "MOIs", inset = 0,
       legend = sapply(MOIs, paste, collapse = " & "), bty = "n")
```

Increasing the number of recurrences without adding data has little effect: the
posterior is close to the value for a single recurrence. 

```{r}
# Allele frequencies (no impact)
fs = list(m1 = c("A" = 0.001, "B" = 0.99)) # Allele frequencies

# The number of recurrences increases but only the first recurrence has data
ys <- list(one_recurrence = list(enroll = list(m1 = "A"), 
                                     recur1 = list(m1 = NA)), 
               two_recurrences = list(enroll = list(m1 = "A"), 
                                      recur1 = list(m1 = NA), 
                                      recur2 = list(m1 = NA)), 
               three_recurrences = list(enroll = list(m1 = "A"), 
                                        recur1 = list(m1 = NA), 
                                        recur2 = list(m1 = NA), 
                                        recur3 = list(m1 = NA)))

# Compute posterior probabilities and extract marginal probabilities
sapply(ys, function(y) {
  suppressMessages(compute_posterior(y, fs, MOIs = c(2,2,rep(1, length(y)-2))))$marg[1,]
})
```

#### Data on more than one episode

When there are data on multiple episodes but no comparable data across episodes,
`compute_posterior` behaves similarly to when data are limited to a single episode:

- returns the prior when episodes with data have MOIs of one 
- remains close to the prior when calls are heteroallelic
- departs from the prior when calls are homoallelic and the allele is rare

```{r}
# Allele frequencies
fs = list(m1 = c("A" = 0.001, "B" = 0.999), 
          m2 = c("A" = 0.001, "B" = 0.999), 
          m3 = c("A" = 0.001, "B" = 0.99))

# Data for different numbers of recurrences
y_hom <- list(enroll = list(m1 = "A", m2 = NA, m3 = NA), 
          recur1 = list(m1 = NA, m2 = "A", m3 = NA), 
          recur2 = list(m1 = NA, m2 = NA, m3 = "A"))

y_het <- list(enroll = list(m1 = c("A", "B"), m2 = NA, m3 = NA), 
          recur1 = list(m1 = NA, m2 = c("A", "B"), m3 = NA), 
          recur2 = list(m1 = NA, m2 = NA, m3 = c("A", "B")))

# Marginal posterior probabilities given MOIs > 1 for episodes without data
suppressMessages(compute_posterior(y_hom, fs))$marg # Prior return 
suppressMessages(compute_posterior(y_het, fs))$marg # Prior proximity
suppressMessages(compute_posterior(y_hom, fs, MOIs = c(2,2,2)))$marg # Departure
```

Data on a single marker can quash the posterior probability of reinfection or
recrudescence. More specifically, one rare match will quash the posterior
probability of reinfection; one mismatch will quash the posterior probability of
recrudescence. However, the output of `compute_posterior` is not the prior over
recurrence states re-weighted to the exclusion of the quashed state. Rather, it
is a transformation of the prior distribution over relationship graphs. This is
demonstrated below: increasing the number of recurrences without adding data
changes the prior distribution over relationship graphs and thus the posterior
probabilities of relapse versus recrudescence (match data), and of relapse
versus reinfection (mismatch data).

```{r}
# Data for different recurrence counts where only the 1st recurrence has data
ys_match <- list(one_recurrence = list(enroll = list(m1 = "A"),
                                       recur_match = list(m1 = "A")),
                 two_recurrences = list(enroll = list(m1 = "A"),
                                        recur_match = list(m1 = "A"),
                                        recur2 = list(m1 = NA)),
                 three_recurrences = list(enroll = list(m1 = "A"),
                                          recur_match = list(m1 = "A"),
                                          recur2 = list(m1 = NA),
                                          recur3 = list(m1 = NA)),
                 four_recurrences = list(enroll = list(m1 = "A"),
                                         recur_match = list(m1 = "A"),
                                         recur2 = list(m1 = NA),
                                         recur3 = list(m1 = NA),
                                         recur4 = list(m1 = NA)))

# Data for different recurrence counts where only the 1st recurrence has data
ys_mismatch <- list(one_recurrence = list(enroll = list(m1 = "A"),
                                          recur_match = list(m1 = "B")),
                    two_recurrences = list(enroll = list(m1 = "A"),
                                           recur_match = list(m1 = "B"),
                                           recur2 = list(m1 = NA)),
                    three_recurrences = list(enroll = list(m1 = "A"),
                                             recur_match = list(m1 = "B"),
                                             recur2 = list(m1 = NA),
                                             recur3 = list(m1 = NA)),
                    four_recurrences = list(enroll = list(m1 = "A"),
                                            recur_match = list(m1 = "B"),
                                            recur2 = list(m1 = NA),
                                            recur3 = list(m1 = NA),
                                            recur4 = list(m1 = NA)))

# Allele frequencies
fs <- list(m1 = c("A" = 0.001, "B" = 0.999))

# Compute posterior probabilities and extract marginal probabilities
results_match <- lapply(ys_match, function(y) suppressMessages(compute_posterior(y, fs)$marg))
results_mismatch <- lapply(ys_mismatch, function(y) suppressMessages(compute_posterior(y, fs)$marg))

# Extract results for the first recurrence
results_match_recur1 <- sapply(results_match, function(result) result[1,])
results_mismatch_recur1 <- sapply(results_mismatch, function(result) result[1,])

# Visualise the change in the marginal probability of the first recurrence
par(mar = c(0,0,0,0))
plot_simplex(c("Recrudescence", "Relapse", "Reinfection"), 0.5) # Plot simplex
xy <- apply(results_match_recur1, 2, project2D) # Project probabilities
points(x = xy["x", ], y = xy["y", ], pch = "—", col = 1:4) # Plot projections
text(x = mean(xy["x", ]), y = mean(xy["y", ]), labels = "match data", pos = 3) # Annotate
xy <- apply(results_mismatch_recur1, 2, project2D) # Project probabilities
points(x = xy["x", ], y = xy["y", ], pch = "|", col = 1:4) # Plot projections
text(x = mean(xy["x", ]), y = mean(xy["y", ]), labels = "mismatch data", pos = 3) # Annotate
legend("left", col = 1:4, pch = "-", pt.cex = 2, bty = "n", legend = 1:4,
       title = "Recurrence\ncount") # legend
```

Increasing MOIs also change the prior distribution over relationship graphs, and
thus the posterior probabilities of relapse versus recrudescence (match data),
and of relapse versus reinfection (mismatch data). However, when MOIs increase,
the amount of data also increases.

```{r}
fs <- list(m1 = c("A" = 0.001, "B" = 0.999))
MOIs <- list(c(1,1), c(2,1), c(3,1), c(2,2), c(3,2), c(3,3), c(4,2))
results_match <- sapply(MOIs, function(x) 
  suppressMessages(compute_posterior(y = ys_match[[1]], fs, MOIs = x)$marg))
results_mismatch <- sapply(MOIs, function(x) 
  suppressMessages(compute_posterior(y = ys_mismatch[[1]], fs, MOIs = x)$marg))
 
par(mar = c(0,0,0,0)) # Visualise the change 
plot_simplex(c("Recrudescence", "Relapse", "Reinfection"), 0.5) # Plot simplex
xy <- apply(results_match, 2, project2D) # Project probabilities
points(x = xy["x", ], y = xy["y", ], pch = 20, col = 1:length(MOIs)) # Plot projections
text(x = median(xy["x", ]), y = mean(xy["y", ]), labels = "match data", pos = 2) # Annotate
xy <- apply(results_mismatch, 2, project2D) # Project probabilities
points(x = xy["x", ], y = xy["y", ], pch = 20, col = 1:length(MOIs)) # Plot projections
text(x = sum(range(xy["x", ]))/2, y = mean(xy["y", ]), labels = "mismatch data", pos = 3) # Annotate
legend("left", col = 1:length(MOIs), pch = 20, pt.cex = 2, bty = "n", 
       legend = gsub("c", "", as.character(MOIs)), title = "MOIs") # legend
```

### Many marker data


## Interpreting uncertainty

Having rejected recrudescence, data on multiple markers are required to reject
reinfection. Having rejected reinfection, data on multiple markers are required
to reject recrudescence. 



The marginal probabilities of relapse / reinfection of different recurrences
with the same data differ slightly because XXX [discuss with YS]



```{r}
suppressMessages(compute_posterior(y = list(enroll = list(m1 = "A"), 
                                            recur1 = list(m1 = "A"), 
                                            recur2 = list(m1 = "A"), 
                                            recur3 = list(m1 = "A")), 
                                   fs = list(m1 = c("A" = 0.05,"B" = 0.95))))$marg
```





## Siblings and relapse
When the model is well specified and recurrent parasites are regular siblings,
the posterior probability of relapse tends to one regardless of the MOIs.

## Clones and recrudescence 
When the model is well specified and recurrent parasites are clones, the
posterior probability of recrudescence tends towards a high but non-certain
probability that depends on the MOIs. 

## Strangers and reinfection 
When the model is well specified and recurrent parasites are strangers, the
posterior probability of reinfection tends towards a high but non-certain
probability that depends on the MOIs. 

## Effect of the prior assumption on relationship graphs 

### Non-recurrent data are informative 

### Effect when the graph grows without data

The graph grows without data as the number of recurrences increases but all data are missing 

### Effect when the graph grows with data

The graph grows without data as the MOI increases



# Misspecified model 

The 




   

