---
title: "Sibling_assumptions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sibling_assumptions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Pv3Rs)
```

# Introduction

Under the Pv3Rs model we make two assumptions that together imply all siblings
are regular full siblings:

1) Siblings are transitive 
2) For each marker, siblings inherit alleles from at most two parents with replacement

More explicitly, to enumerate all relationship graphs, we enumerate all sibling
partitions over each clonal partition, generating graphs within which sibling
(and clonal) components are fully connected (assumption one). When enumerating
all IBD partitions consistent with a given relationship graph, we assume each
sibling cell can be split into one or two IBD clusters (assumption two).[Yong
See, am I missing anything?]

Consequently, the Pv3Rs model is misspecified when data come from

- parent-child-like siblings (not transitive),
- half siblings (inherit from three parents),
- mieotic siblings (not independent),

all of which could be present in a brood because gametocytes can self (necessary
for parent-child-like siblings), sporozoites can come from different oocysts
(necessary for parent-child-like siblings and for half siblings), and
sporozoites can come from a single oocyst (necessary for mieotic siblings); see
Figure XXX.

We aim to relax assumptions one and two. In the meantime, we describe how the
model behaves when it is misspecifed because data come from parent-child-like
siblings, half siblings and meiotic siblings. For these edge cases, a full
analysis of the model's behaviour is beyond our current scope.

# Misspecification due to parent-child-like siblings

# Misspecification due to half siblings

# Misspecification due to mieotic siblings


