---
title: Understand prior impact
description: Exlores the impact of the prior in a very limited setting
---

# Summary

In a very limited setting (single recurrence, MOIs of one or two per episode,
extreme and contrived data and allele frequencies), the following two questions
are explored.

1) What is the impact of the prior? 
2) Can the prior offset misspecification?

In summary: the prior does impact the posterior and can offset misspecification,
but not always. More specifically, 

1) A high-relapse prior can generate a high-relapse posterior despite data
consistent with reinfection.
2) A high-relapse prior can rescue the relapse posterior when data are
consistent with siblings but break the assumption that siblings draw their
alleles from at most two parental gametes.
3) A high-recrudescence prior cannot rescue the recrudescence posterior when
data are consistent with clones with a single miss-match, e.g., due to
genotyping error.

For data consistent with relapse, the results are more nuanced. In summary, the
posterior transitions between probable relapse and probable reinfection with the
increasing frequency of the allele that matches across episodes. When the Pv3Rs
model is fit to only 10 markers, the transition is gradual, and a low-relapse
prior can lower the relapse posterior when it is otherwise high. When the Pv3Rs
model is fit to 100 markers, the transition is is abrupt. A low-relapse prior
can generate a low-relapse posterior when it is otherwise high, but only within
a narrow window of frequency (this window might differ for more realistic data). 


```{r}
library(Pv3Rs)
```

# Some contrived examples on which insight is based

## Generate synthetic data, frequencies and priors 

```{r}
marker_count <- 100 # Number of markers
ms <- paste0("m", 1:marker_count) # Marker names

# Generate per-episode data 
all_As <- sapply(ms, function(t) "A", simplify = F)
all_Bs <- sapply(ms, function(t) "B", simplify = F)
BAAAAA <- all_As; BAAAAA[["m1"]] <- "B"
ABABAB <- sapply(ms, function(t) ifelse(gtools::odd(as.numeric(gsub("m","",t))), "A", "B"), simplify = F)
BC_BABAB <- ABABAB; BC_BABAB[["m1"]] <- c("B", "C") # Three alleles at m1

# Generate paired-episode data 
sib <- list(enrol = all_As, recur = ABABAB) # Perfect mosaic
str <- list(enrol = all_As, recur = all_Bs) # Perfect mismatch
mis_L <- list(enrol = all_As, recur = BC_BABAB) # Perfect mosaic with draw from three alleles at m1
mis_C <- list(enrol = all_As, recur = BAAAAA) # Perfect mosaic with draw from three alleles at m1

# Generate frequencies
f_rare <- 0.1 # Frequency of rare allele
fs_rareA <- c(list(m1 = c("A" = 1/3, "B" = 1/3, "C" = 1/3)),  
              sapply(ms[-1], function(m) c("A" = f_rare, "B" = 1 - f_rare), simplify = FALSE))

# Specify priors
prior_hi_L <- array(c(0, 0.99, 0.01), dim = c(1,3), dimnames = list(NULL, c("C", "L", "I")))
prior_hi_I <- array(c(0, 0.01, 0.99), dim = c(1,3), dimnames = list(NULL, c("C", "L", "I")))
prior_hi_C <- array(c(0.95, 0.025, 0.025), dim = c(1,3), dimnames = list(NULL, c("C", "L", "I")))
```

## Plot sythetic data (first 10 of 100 markers)
```{r echo = F, out.width="100%"}
# Plot the data on the first 10 makers
plot_data(list(sib = sapply(sib, head, 10, simplify = F), 
               str = sapply(str, head, 10, simplify = F), 
               mis_L = sapply(mis_L, head, 10, simplify = F),
               mis_C = sapply(mis_C, head, 10, simplify = F)),
          fs = fs_rareA[1:10])
```

## Compare posterior with and without non-uniform priors

### High-relapse prior increases relapse posterior when data are consistent with reinfection: 
```{r}
suppressMessages(compute_posterior(str, fs_rareA))$marg
suppressMessages(compute_posterior(str, fs_rareA, prior_hi_L))$marg
```

### High-relapse prior rescues relapse posterior when data are consistent with half-sib misspecification:
```{r}
suppressMessages(compute_posterior(mis_L, fs_rareA))$marg
suppressMessages(compute_posterior(mis_L, fs_rareA, prior_hi_L))$marg
```

### High-recrudescence prior cannot rescue recrudescence posterior given mismatched data: 
```{r}
suppressMessages(compute_posterior(mis_C, fs_rareA))$marg
suppressMessages(compute_posterior(mis_C, fs_rareA, prior_hi_C))$marg
```

### High-reinfection prior sometimes impacts an otherwise high-relapse posterior:

High-reinfection prior lowers the posterior probability of relapse when the data
are consistent with relapse and when there are 10 markers (blue lines). When
there are 100 markers (pink lines) the prior has an impact in the vicinity of
the frequency leading to abrupt change between probable relapse and probable
reinfection. The allele that matches has to have a very high frequency in the
case of data that are almost consistent with recrudescence.

```{r echo = FALSE, out.width="100%"}
par(mfrow = c(1,2))
for(case in c("sib", "mis_C")) {
  
  fs <- seq(f_rare, 1, 0.01); names(fs) <- fs
  if(case == "sib") {
    y <- sib
    title_ <- "50% match"
  } else {
    y <- mis_C
    title_ <- "99% match"
  }
  
  results <- sapply(fs, function(f_comn){
    
    fs_comnA <- c(list(m1 = c("A" = 1/3, "B" = 1/3, "C" = 1/3)),  
                  sapply(ms[-1], function(m) c("A" = f_comn, "B" = 1 - f_comn), simplify = FALSE))
    
    # Posterior relapse probabilities
    c(UniL_100 = suppressMessages(compute_posterior(y, fs_comnA))$marg[,"L"],
      LowL_100 = suppressMessages(compute_posterior(y, fs_comnA, prior_hi_I))$marg[,"L"],
      UniL_10 = suppressMessages(compute_posterior(sapply(y, head, 10, simplify = F), 
                                                   fs_comnA[1:10]))$marg[,"L"],
      LowL_10 = suppressMessages(compute_posterior(sapply(y, head, 10, simplify = F),
                                                   fs_comnA[1:10], prior_hi_I))$marg[,"L"])
  })
  
  plot(NULL, xlim = range(fs), ylim = c(0,1), bty = "n", main = title_, 
       ylab = "Posterior relapse", xlab = "Allele frequency")
  
  lines(y = results["UniL_100",], x = fs, col = "hotpink")
  lines(y = results["LowL_100",], x = fs, col = "hotpink", lty = "dashed")
  lines(y = results["UniL_10",], x = fs, col = "blue")
  lines(y = results["LowL_10",], x = fs, col = "blue", lty = "dashed")
  
  if(case == "sib"){
    legend("topright", legend = c("Uniform prior", "Low relapse prior"), 
           lty = c("solid", "dashed"), bty = "n")
    legend("right", legend = c("100 markers", "10 markers"), 
           col = c("hotpink", "blue"), bty = "n", pch = 15)    
  }
}
```

# Rescuing misspecification 

The simulated data used to demonstrate the impact of half sibling misspecification in 
[Understand posterior probabilities](https://aimeertaylor.github.io/Pv3Rs/articles/posterior-probabilities.html#half)
were re-analysed using a high-relapse prior (0.9 relapse), recovering high-relapse posteriors: 

```{r echo = FALSE, out.width="100%"}
#===============================================================================
# Magic numbers / quantities
#===============================================================================
n_repeats <- 10 # Number of simulations per parameter combination
max_n_markers <- 150 # 150 # Number of markers for which RG likelihood returned
c_cutoff <- 99 # Switch from Dirichlet r.v. to 1/n_alleles above c_cutoff
seed <- 10 # For reproducibility
prior <- array(c(0.05, 0.9, 0.05), dim = c(1,3), dimnames = list(NULL, c("C", "L", "I")))
generate_results <- FALSE

#===============================================================================
# Stores for results
#===============================================================================
ps_store <- list() # p for posterior
ps_store_all_ms_uniform <- list() # all_ms for all marker counts
ps_store_all_ms_admix_rare <- list() # for admixed case with some rare freqs

#===============================================================================
# Load frequencies and data
#===============================================================================
load("RelationshipStudy/output_HalfSib.PCLikeSib.rda")
fs_store <- output_HalfSib.PCLikeSib$Half$fs_store
ys_store <- output_HalfSib.PCLikeSib$Half$ys_store
m_rorder <- output_HalfSib.PCLikeSib$Half$m_rorder

if (generate_results) {
  
  #=============================================================================
  # Generate results for markers 1:max_n_markers
  #=============================================================================
  c <- 100 # For uniform allele frequencies and all markers
  admixture <- FALSE # For parents from the same population
  fs <- fs_store[[as.character(c)]] # Extract frequencies
  for(i in 1:n_repeats){
    y_all_markers <- ys_store[[as.character(c)]][[sprintf("admixture_%s", admixture)]][[i]]
    for(j in 1:max_n_markers){
      marker_subset <- m_rorder[1:j]
      y <- sapply(y_all_markers, function(x) x[marker_subset], simplify = FALSE)
      ps <- suppressMessages(compute_posterior(y, fs, prior))
      ps_store_all_ms_uniform[[as.character(i)]][[j]] <- ps$marg
    }
  }
  
  c <- 0.5 # For admixed case with imbalanced allele frequencies
  admixture <- TRUE # For parents from different populations
  fs <- fs_store[[as.character(c)]] # Extract frequencies
  for(i in 1:n_repeats){
    y_all_markers <- ys_store[[as.character(c)]][[sprintf("admixture_%s", admixture)]][[i]]
    for(j in 1:max_n_markers){
      marker_subset <- m_rorder[1:j]
      y <- sapply(y_all_markers, function(x) x[marker_subset], simplify = FALSE)
      ps <- suppressMessages(compute_posterior(y, fs, prior))
      ps_store_all_ms_admix_rare[[as.character(i)]][[j]] <- ps$marg
    }
  }
  
  #=============================================================================
  # Bundle magic numbers, data and results
  #=============================================================================
  output <- list(
    n_repeats = n_repeats,
    max_n_markers = max_n_markers,
    m_rorder = m_rorder,
    fs_store = fs_store,
    c_cutoff = c_cutoff,
    seed = seed,
    prior = prior,
    ps_store = ps_store,
    ps_store_all_ms_uniform = ps_store_all_ms_uniform,
    ps_store_all_ms_admix_rare = ps_store_all_ms_admix_rare)
  
  save(output, file = "halfsib_rescue.rda")
} else {
  load("halfsib_rescue.rda")
}



#=============================================================================
# Results
#=============================================================================
cols <- RColorBrewer::brewer.pal(n = output$n_repeats, "Paired") # For repeats
par(mfrow = c(2,1))

# Compute effective cardinality cumulatively in the order of markers genotyped
cum_card_eff <- sapply(output$fs_store, function(fs) {
  cumsum(sapply(fs[output$m_rorder], function(x) 1/sum(x^2)))})

# Determine which fs in cum_card_eff are equifrequent
equifs <- which(as.numeric(colnames(cum_card_eff)) > output$c_cutoff)
```

```{r echo = FALSE}
# ------------------------------------------------------------------------------
# Plot the posterior relapse probability trajectories
# ------------------------------------------------------------------------------
plot(NULL, bty = "n", las = 1, xaxt = "n", xlim = c(1,max_n_markers),
     xlab = "Marker count (effective cardinality)",
     ylim = c(0,1), ylab = "Posterior relapse probability")
legend("right", lwd = 2, col = cols, legend = 1:output$n_repeats, bty = "n",
       cex = 0.5, title = "Repeat")

# Add horizontal axis
axis_at <- c(1, seq(0, max_n_markers, 50)[-1])
axis(side = 1, at = axis_at, cex.axis = 0.7) # Marker count
axis(side = 1, line = 1, at = axis_at, cex.axis = 0.6, # Effective cardinality
     tick = F, labels = sprintf("(%s)", cum_card_eff[,equifs][axis_at]))

# Add trajectories
for(i in 1:output$n_repeats){
  lines(x = 1:max_n_markers,
        y = sapply(output$ps_store_all_ms_uniform[[as.character(i)]],
                   function(x) x[,"L"]),
        col = cols[i], lwd = 2)
}
```

```{r echo = FALSE}
# ------------------------------------------------------------------------------
# Plot the posterior relapse probability trajectories
# ------------------------------------------------------------------------------
plot(NULL, xlim = c(1,max_n_markers), ylim = c(0,1), bty = "n", las = 1,
     xaxt = "n", xlab = "Marker count (effective cardinality)",
     ylab = "Posterior relapse probability")

legend("right", col = cols, lwd = 2, inset = 0.1, legend = (1:output$n_repeats)-1,
       bty = "n", cex = 0.5, title = "Repeats")

# Add horizontal axis
axis(side = 1, at = axis_at, cex.axis = 0.7) # Marker count
axis(side = 1, line = 1, at = axis_at, cex.axis = 0.6, # Effective cardinality
     tick = F, labels = sprintf("(%s)", round(cum_card_eff[,"0.5"][axis_at])))

# Add trajectories
for(i in 1:output$n_repeats){
  y <- sapply(output$ps_store_all_ms_admix_rare[[as.character(i)]],
              function(x) x[,"L"])
  lines(x = 1:max_n_markers, y = y, col = cols[i], lwd = 2)
}
```
